Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
former, but unfortunately, sliding still fails:
\(
  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
\),
since the left-hand side will emit the pomset \([x := a, y := b]\) while the
right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
behaviour, we'd need to have a significantly more sophisticated model of
composition which effectively takes nontermination into account, which we will
leave to future work.

@@

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! The semantics of reads and writes are then given
by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
\cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}

@@

>>>>

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f_0 || f_1 = \lambda (a_0, a_1). 
  & \{\iota_0 ((b_0, b_1), \alpha_0 || \alpha_1) 
    \mid \iota_0 (b_i, \alpha_i) \in f_i\;a_i\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0) \land \iota_1 \alpha_1 \in f_1\;a_1\} 
  \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0 \land \iota_0 (a_1, \alpha_1) \in f_1\;a_1\}: \ms{Set}_{\ms{PO}}(A \otimes A', B \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
former, but unfortunately, sliding still fails:
\(
  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
\),
since the left-hand side will emit the pomset \([x := a, y := b]\) while the
right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
behaviour, we'd need to have a significantly more sophisticated model of
composition which effectively takes nontermination into account, which we will
leave to future work.

@@

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! The semantics of reads and writes are then given
by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
\cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}

@@

We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}. Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
\begin{equation}
  \begin{aligned}
    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
    \end{aligned}
\end{equation}
